Dynamic Stochastic Approximation Monte Carlo

### 2018-06-28 papers/histogram/sad.tex
added citations and addressed problems in sad paper.

The organization of this paper is as follows: In Section II, we
describe in detail several broad-histogram methods used to calculate
the density of states for the liquid-vapor system.  Section III
contains the results and comparison of the methods applied to the
square-well fluid.  In Section IV, we discuss the performance of the
various methods applied to the model system.

\davidsays{Fix this, it was moved when I reorganized the methods
  section} In this work, We compare a variety of broad histogram
methods.  We outline the general workings of each algorithm that we
developed in detail while summarizing algorithms that originated
elsewhere.  The following methods are introduced and applied to the
square-well fluid: Transition Matrix Monte-Carlo (TMMC), Wang-Landau
(WL), Wang-Landau Transition Matrix Monte-Carlo (WL-TMMC), Stochastic
Approximation Monte-Carlo (SAMC), and Dynamic Stochastic Approximation
(SAD).

\section{Broad histogram methods}

@@

\davidsays{We should discuss continuous energies and binning somewhere
  here...}

\section{Weight-based methods}

We will begin by introducing two related weight-based methods which
rely on a weight function $w(E)$, noting that the method proposed in
this paper (discussed in Section~\ref{sec:sad}) also falls in this
category.  In these algorithms, the probability of accepting a move is
given by
\begin{equation}
	\mathcal{P}(\epsilon_\text{old} \rightarrow \epsilon_\text{new})
	= \min\left[1,\frac{w(\epsilon_\text{old})}{w(\epsilon_\text{new})}\right]
\end{equation}
This biases the simulation in favor of energies with low weights.  A
set of weights that are proportional to the density of states of the
system $D(\epsilon)$ will result in an entirely flat histogram.  Thus
the goal of each of these methods is to achieve a flat histogram.  For
numerical reasons, it is common to store and work with the natural
logarithm of the weights, since the weights may vary over more than a
few hundred orders of magnitude.  Note that the logarithm of the
density of states is the entropy at that energy, so we can think of
the logarithm of the weights as an approximation of the entropy
\davidsays{make this less colloquial}.

Each weighted density approach use a random walk in energy space to
estimate the density of states.  The core of these approaches
is to continuously update the weights at each step of the simulation
\begin{equation}
	\ln{w_{t+1}(\epsilon)}=\ln{w_{t}(\epsilon)}
	+\gamma
\end{equation}
where $t$ is number of the current move, and $\gamma$ is an update
factor.  This update causes the random walk to avoid energies that
have been frequently sampled, leading to a rapid exploration of energy
space.  This approach, however, violates detailed balance, due to the
acceptance probabilities changing with each move.  The severity of
this violation decreased as we decrease $\gamma$.  The various
weight-based methods discussed here differ in how the schedule the
decreasing of $\gamma$.

@@

Wang-Landau's approach~\cite{wang2001efficient,wang2001determining,
  landau2014guide} begins with $\gamma=1$, and then decreases $\gamma$
in discrete stages.  We track the number of steps at each energy during
each stage in a histogram.  When that histogram is sufficiently flat,
$\gamma$ is decreased by a specified factor, which is usually
$\frac12$.  The flatness is defined by the ratio between the minimum
value of the histogram and its average value.  When this flatness
reaches a specified threshold (typically 0.8), the $\gamma$ value is
decreased.  This approach requires that the energy range of interest
be known in advance, and difficulties can occur with this flatness
criteria due to the fact that some energies in this energy range might
never be sampled~\cite{haber2014transition}.  The entire process is
repeated until $\gamma$ reaches a desired cutoff is reached.

@@

\davidsays{We need to look up these references.}  While this approach
is very efficient and has been widely uses, it suffers a few
shortcomings.  Firstly, the set of energies must be specified, which
may require multiple simulations.  Secondly, while Wang-Landau
converges quickly, it does not in general converge to the true density
of states.  It can and does decrease $\gamma$ so quickly that it will
never (for any cutoff value) decrease the error in the density of
states beyond a given value.

\subsection{SAMC}
The Stochastic Approximation Monte Carlo (SAMC) algorithm addresses
the lack of convergence of Wang-Landau's approach with a simple
schedule by which the update factor $\gamma$ is continuously
decreased~\cite{liang2007stochastic, werlich2015stochastic,
  schneider2017convergence}.  The update factor is defined in the
original implementation~\cite{liang2007stochastic} in terms of an
arbitrary tunable parameter $t_0$ and a scale factor
$\gamma_0$. \davidsays{Did it originally have a scale factor?}
\begin{align}
\gamma_{t}^{\text{SA}} =\gamma_0 \frac{t_0}{\max(t_0,t)}
\end{align}
where as above $t$ is the number of moves that have been attempted.
The major advantage that SAMC offers is its proven convergence.
Provided the update factor satisfies
\begin{align}
\sum_{t=1}^\infty \gamma_{t} = \infty \quad\textrm{and}\quad
\sum_{t=1}^\infty \gamma_{t}^\zeta < \infty
\end{align}
where $\zeta \in \{1,2\}$, Liang has shown that the weights are proven
to converge to the true density of states~\cite{liang2006theory,
  liang2007stochastic}.  In addition, the energy range need not be
known a priori.  The time to converge depends only on the choice of
parameters $t_0$ and $\gamma_0$.  Unfortunately, there is no
prescription for finding acceptable values for these parameters, and
while the algorithm formally converges, for a poor choice of $t_0$
that convergence may be far too slow to be practical.  This work
proposes a means for dynamically setting the update factor $\gamma_t$
that should robustly converge on a variety of systems.

\section{Transition-matrix methods}

@@

\section{Results}

>>>>

The organization of this paper is as follows: In Section~\ref{sec:histogram}, we
describe in detail broad-histogram methods.  Section~\ref{sec:weight} outlines
weight-based histogram methods used in this work. Section~\ref{sec:transition}
details the workings of the transition-matrix methods studied in this work.
Section~\ref{sec:sad} introduces the dynamic stochastic approximation method.
In Section~\ref{sec:results}, we discuss the performance of the various methods 
applied to the model system.

In this work, We compare a variety of broad histogram
methods.  We outline the general workings of each algorithm that we
developed in detail while summarizing algorithms that were developed in other
works.  The following methods are introduced and applied to the
square-well fluid: Transition Matrix Monte-Carlo (TMMC), Wang-Landau
(WL), Wang-Landau Transition Matrix Monte-Carlo (WL-TMMC), Stochastic
Approximation Monte-Carlo (SAMC), and Dynamic Stochastic Approximation
(SAD).

\section{Broad histogram methods}\label{sec:histogram}

@@

\davidsays{We should discuss continuous energies and binning somewhere
  here...}
\jpsays{I put some text about this but it was under a sad subsection also
may need reworked and/or edited.}

\section{Weight-based methods}\label{sec:weight}

We will begin by introducing two related weight-based methods which
rely on a weight function $w(E)$, noting that the method proposed in
this paper (discussed in Section~\ref{sec:sad}) also falls in this
category.  In these algorithms, the probability of accepting a move is
given by
\begin{equation}
	\mathcal{P}(\epsilon_\text{old} \rightarrow \epsilon_\text{new})
	= \min\left[1,\frac{w(\epsilon_\text{old})}{w(\epsilon_\text{new})}\right]
\end{equation}
the probability biases the simulation in favor of energies with low weights.  
A set of weights that are proportional to the density of states of the
system $D(\epsilon)$ will result in an entirely flat histogram.  Thus
flat histogram is a criteria for convergence for these methods.  To avoid 
overflow error , since the weights may vary over more than a
few hundred orders of magnitude, the natural logarithm of the weights are used 
in this work.  Since $D(\epsilon) = S(\epsilon)$, the logarithm of the weights 
can be thought of as an approximation of the entropy.

Each weighted density approach uses a random walk in energy space to
estimate the density of states.  The core of these approaches
is to continuously update the weights at each step of the simulation
\begin{equation}
	\ln{w_{t+1}(\epsilon)}=\ln{w_{t}(\epsilon)}
	+\gamma
\end{equation}
where $t$ is number of the current move, and $\gamma$ is an update
factor.  This update causes the random walk to avoid energies that
have been frequently sampled, leading to a rapid exploration of energy
space.  This approach, however, violates detailed balance, due to the
acceptance probabilities changing with each move.  The severity of
this violation decreased as we decrease $\gamma$.  The various
weight-based methods discussed here differ in how the decreasing of $\gamma$
is scheduled.

@@

Wang-Landau's approach~\cite{wang2001efficient,wang2001determining,
  landau2014guide} begins with $\gamma=1$, and then decreases $\gamma$
in discrete stages.  We track the number of steps at each energy during
each stage in a histogram.  When that histogram is sufficiently flat,
$\gamma$ is decreased by a specified factor, which is usually
$\frac12$.  The flatness is defined by the ratio between the minimum
value of the histogram and its average value.  When this flatness
reaches a specified threshold (typically 0.8), the $\gamma$ value is
decreased.  This approach requires that the energy range of interest
be known in advance, and difficulties can occur with this flatness
criteria due to the fact that some energies in this energy range might
never be sampled~\cite{haber2014transition}.  The entire process is
repeated until $\gamma$ reaches a desired cutoff.

@@

While this approach
is very efficient and has been widely used, it suffers a few
shortcomings.  Firstly, the set of energies must be specified
~\cite{wang2001efficient, schulz2003avoiding, yan2003fast}, which may 
require multiple simulations. Secondly, while Wang-Landau converges quickly, 
it does not in general converge to the true density of states
~\cite{belardinelli2008analysis, zhou2008optimal}.  It can and does decrease
$\gamma$ so quickly that it will never (for any cutoff value) decrease the 
error in the density of states beyond a given value.

\subsection{SAMC}
The Stochastic Approximation Monte Carlo (SAMC) algorithm addresses
the lack of convergence of Wang-Landau's approach with a simple
schedule by which the update factor $\gamma$ is continuously
decreased~\cite{liang2007stochastic, werlich2015stochastic,
  schneider2017convergence}.  The update factor is defined in the
original implementation~\cite{liang2007stochastic} in terms of an
arbitrary tunable parameter $t_0$ and in future works~\cite{werlich2015stochastic} with a scale factor
$\gamma_0$.
\begin{align}
\gamma_{t}^{\text{SA}} =\gamma_0 \frac{t_0}{\max(t_0,t)}
\end{align}
where as above $t$ is the number of moves that have been attempted.
The major advantage that SAMC offers is its proven convergence.
Provided the update factor satisfies
\begin{align}
\sum_{t=1}^\infty \gamma_{t} = \infty \quad\textrm{and}\quad
\sum_{t=1}^\infty \gamma_{t}^\zeta < \infty
\end{align}
where $\zeta \in \{1,2\}$, Liang has shown that the weights are proven
to converge to the true density of states~\cite{liang2006theory,
  liang2007stochastic}.  In addition, the energy range need not be
known a priori.  The time to converge depends only on the choice of
parameters $t_0$ and $\gamma_0$.  Unfortunately, there is no
prescription for finding acceptable values for these parameters, and
while the algorithm formally converges, for a poor choice of $t_0$
that convergence may be far too slow to be practical.  This work
proposes a means for dynamically setting the update factor $\gamma_t$
that should robustly converge on a variety of systems.

\section{Transition-matrix methods}\label{sec:transition}

@@

\section{Results}\label{sec:results}
